Automated Analysis of Muscle X-ray Diffraction \\ Imaging with MCMC

### 2015-05-29 paper.tex
Incorporated Tom's edits, removed passive voice

% Figures
\newcommand{\figureimgs}{
\begin{figure}[tbp]
  \centering
  \includegraphics[width=\linewidth]{figures/x_ray_image_montage}
  \vspace{-12pt}
  \caption{\label{fig:imgs}
  	X-ray diffraction images.
    A) An X-ray diffraction image exhibits $d_{10}$, $d_{11}$, and
    $d_{20}$ peaks; the first and last are proportional to the
    distance between adjacent rows of thick filaments. The $d_{10}$ is
    the primary peak being targeted for automated detection and
    fitting. B) In this image, two fibers at slight angles to each
    other have generated multiple $d_{10}$ peaks which must be
    segregated during the detection process.  
	}
	\vspace{-2pt}
\end{figure}
}

\newcommand{\figureworkflow}{
\begin{figure}[tbp]
  \centering
  \includegraphics[width=\linewidth]{figures/img_analysis}
  \vspace{-8pt}
  \caption{\label{fig:workflow}
  	Analysis workflow.
    The blocked center region is detected and used for registration.
    Low signal regions and edges are masked. The image is smoothed and
    peaks are detected, clustered, and classified. The background
    (minus that surrounding diffraction lines) is computed and
    subtracted. Regions of interest surrounding the $d_{10}$ peaks are
    extracted and the generating distributions are fit with an MCMC
    sampler. We marginalize across peak parameters other than those of
    interest.  
	}
	\vspace{-2pt}
\end{figure}
}

\begin{document}
\title{Automated Analysis of Muscle X-ray Diffraction \\ Imaging with MCMC}
\author{C.\ David Williams$^{1,2}$,  Magdalena Balazinska$^2$, Thomas Daniel$^1$ \\
\affaddr{$^1$Department of Biology, $^2$Department of Computer Science \& Engineering}\\
\affaddr{University of Washington} \\
}
\maketitle

@@

Biologists have studied muscle structure and dynamics since antiquity
but it is only with the advent of X-ray diffraction techniques in the
1960s that it became possible to visualize the orientation and spacing
of the semicrystalline proteins that generate and transmit force
within muscle \cite{Millman1998}. In X-ray diffraction, structural
information is recorded when an X-ray beam passes through and scatters
off a muscle sample to produce an image such as that shown in
Figure~\ref{fig:imgs}.

Today, human experts manually extract structural parameters
from X-ray images using NIH ImageJ. This produces repeatable
measurements but remains subjective, fails to provide confidence
intervals for the measured values, and is not reproducible by naive
digitizers. Additionally hand digitization is a time-intensive
analytic technique, with a single digitizer able to process only a few
hundred images a day.  This has historically been sufficient, but new
high-speed imaging systems are allowing the investigation of
short-timescale components of muscle contraction and generating data
sets with many thousands of images. The need for an automated and
reproducible image analysis tool chain is clear.

The goal of our project is to build a service for the automated
analysis of X-ray images of muscle structures.  Users should specify
the analysis that they need using a declarative query interface and
the system should automatically process the user's image database.  In
this extended abstract, we present the first components of the
processing tool chain at the heart of this service. The toolchain
first segments the diffraction image into regions of interest using
conserved features and then samples the possible parameter values with
a Markov chain Monte Carlo approach.

@@

We initially focus on measuring the $d_{10}$ parameter, a crucial
spacing in muscle shown in Figure~\ref{fig:imgs}. The $d_{10}$ spacing
determines the distance which muscle's molecular motors must bridge in
order to bind and generate force \cite{Williams2013}. This distance
changes during contraction, regulating the force produced.

Images generated during experiments share several key features, which
serve as challenges or fiduciary marks during analysis. As seen in
Figure~\ref{fig:imgs}, the brightest part of the image background is
occluded by a circular stop. This physical block is put in place to
prevent damage to the detector from the high photon flux seen at the
center of the X-ray beam. Surrounding the blocked region, the
remainder of the image displays an exponentially decaying background.
Interrupting this background are symmetric pairs of diffraction peaks
we locate and model.  Our core data analysis pipeline includes two
steps: image segmentation and image modeling with MCMC processes.

@@

The dark central circular blocked region is first located and acts as
a relative reference point for subsequent operations. Consistent with
experimental design, it is assumed to contain the center of the
diffraction pattern, although the center of the diffraction pattern is
not assumed to be at the center of the blocked region. The edge of the
background surrounding the block is the brightest region, so the image
is first split between areas with values less than and greater than
two standard deviations above the mean. This partitioning yields a
binary image where the center blocked region is surrounded by a halo
of the upper end of the pattern background and occasional dots where
diffraction peaks rise more than two standard deviations above
background. This binary image is converted to a hierarchical contour
set with OpenCV. The blocked region is taken as the inner-most contour
and then modeled as the smallest enclosing circle, shown as dark
magenta in Figure~\ref{fig:workflow}.

\dave{I cut this down and simplified. It can be removed or further
reworked if desired.}
With the central blocked region located, we identify local maxima in 3
by 3 pixel groups after Gaussian smoothing and mask those found in
regions unlikely to provide peaks of interest.  Resulting maxima are
accepted or rejected as peaks based on masking parameters.  Masked
rejection-regions consist of: 1) a circular zone around the central
blocked region where the blocking generates non-peak local maxima, 2)
areas below the 80th percentile where detector noise dominates, 3)
areas near the image edge where peaks are partially cropped. The
resulting unmasked area from which we keep maxima is depicted as light
purple overlay in Figure~\ref{fig:workflow}.

@@

Next, maxima are clustered into peak pairs based on their distance
and angle from the center of the blocked region. Starting with those
maxima nearest the blocked region, a corresponding maxima is sought an
equal distance away from the blocked region and located such that the
angle formed by the two maxima and the center of the blocked region is
180$^\circ$. If no matched maxima exists within a 10\% tolerance, the
maxima is discarded as a spurious peak. With peak pairs now identified
(shown as color matched dots in Figure~\ref{fig:workflow}), the
diffraction center is identified by taking the mean location between
peak pairs and the background is fit and subtracted.

The background is subtracted by first masking arcs encompassing peak
pairs and then fitting a double exponential to a radial profile of the
remaining image. An arc swept 12$^\circ$ out on either side of each
peak pair masks the effect of the diffraction peaks on the background
(shown as a light gray arc under the peak pairs in
Figure~\ref{fig:workflow}). We calculate a radial profile of the image
around the pattern center, omitting diffraction line regions. We fit a
double exponential function of the form $background = a+ b e^{-x c} +
d e^{-x e}$ to the radial profile. From these parameters we generate
an estimated background image and subtract it from the real image,
allowing us to extract the $d{10}$ peaks as regions of interest (ROIs)
unhindered by an overlaid diffraction background.

@@

With the background subtracted and the $d_{10}$ peaks identified and
isolated from the rest of the image as ROIs, we apply Markov chain
Monte Carlo (MCMC) sampling to determine the probability distributions
from which the peak parameters could be drawn. We treat the peaks as
being drawn from an underlying Pearson VII distribution, commonly used
to fit X-ray diffraction peaks. This process allows us to generate
possible peak matches using five parameters: peak center x-location,
peak center y-location, peak height, peak spread, and peak decay rate.
We perform an initial peak fitting by residual minimization between a
generated peak and the extracted ROI. This gives us a set of starting
parameters that we use, with random variation, to initialize the
positions of the MCMC agents that will explore the model space.

@@

We apply our workflow to a test corpus of 1,220 images generated using
X-ray diffraction during insect flight muscle research at Argonne
National Laboratory's BioCAT Beamline. Sample high-quality and
challenging images are shown in Figures~\ref{fig:imgs}A and
\ref{fig:imgs}B. Our technique segments these images into regions of
interest, identifies the approximate locations of the diffraction
peaks, subtracts background from the regions surrounding the peaks,
and computes the distribution of possible parameters which underly the
shown peaks.  We find that our initial image segmentation step
successfully models the center blocked region of each image in greater
than 99\% of images in our test corpus of 1,220 images. The overall
process allows us to calculate peak-to-peak distances to sub-pixel
accuracy with a confidence interval of 90\%. MCMC sampling combined
with image segmentation allows us to precisely, accurately, and
automatically locate the $d_{10}$ peak center and thus calculate the
lattice spacing measured by a diffraction image to within 0.03 nm.

@@

>>>>

% Figures
\newcommand{\figureimgs}{
\begin{figure}[tbp]
  \centering
  \includegraphics[width=\linewidth]{figures/x_ray_image_montage}
  \vspace{-12pt}
  \caption{\label{fig:imgs}
  	X-ray diffraction images.
    A) An X-ray diffraction image exhibits $d_{10}$, $d_{11}$, and
    $d_{20}$ peaks; the first and last are proportional to the
    distance between adjacent rows of thick filaments. B) In this
    image, two fibers at slight angles to each other have generated
    multiple $d_{10}$ peaks which must be segregated during the
    detection process.  
	}
	\vspace{-2pt}
\end{figure}
}

\newcommand{\figureworkflow}{
\begin{figure}[tbp]
  \centering
  \includegraphics[width=\linewidth]{figures/img_analysis}
  \vspace{-8pt}
  \caption{\label{fig:workflow}
  	Analysis workflow.
    The blocked center region is detected and used as a landmark.  Low
    signal regions and edges are masked. The image is smoothed and
    peaks are detected, clustered, and classified. The background
    (sans that surrounding diffraction lines) is computed and
    subtracted. Regions of interest surrounding the $d_{10}$ peaks are
    extracted and the generating distributions are fit with an MCMC
    sampler. We marginalize across peak parameters other than those of
    interest.  
	}
	\vspace{-2pt}
\end{figure}
}

\begin{document}
\title{Automated Analysis of Muscle X-ray Diffraction \\ Imaging with MCMC}
\author{C.\ David Williams$^{1}$,  Magdalena Balazinska$^2$, Thomas Daniel$^1$ \\
\affaddr{$^1$Department of Biology, $^2$Department of Computer Science \& Engineering}\\
\affaddr{University of Washington} \\
}
\maketitle

@@

The molecular regulation of force generating protein systems remains a
fundamental open problem in biology. Muscle uses a highly organized
lattice of interacting elastic and force generating molecules to
create controlled macro-scale movement \cite{Millman1998}. The advent
of advanced X-ray diffraction imaging lets us analyze protein motions
at spatial and temporal scales never previously realized, giving novel
insight into the molecular basis of motion in living systems.

Surprisingly, we still use human experts to manually extract
structural parameters from X-ray images with NIH ImageJ. This produces
repeatable measurements but is subjective, fails to provide confidence
intervals for the measured values, and is not reproducible by naive
digitizers. Additionally hand digitization is a time-intensive
analytic technique, with a single digitizer able to process only a few
hundred images a day.  Historically this rate has been sufficient, but
new high-speed imaging systems let us investigate short-timescale
components of muscle contraction and generate data sets with many
thousands of images. The need for an automated and reproducible image
analysis tool chain is clear.

We seek to build a service for the automated analysis of muscle
structure X-ray images. Users should specify the analysis they need
using a declarative query interface and the system should
automatically process the user's image database. In this extended
abstract, we present the first components of the processing tool chain
at the heart of this service. The toolchain first segments the
diffraction image into regions of interest using conserved features
and then samples the possible parameter values with a Markov chain
Monte Carlo approach.

@@

We initially focus on the $d_{10}$ parameter that determines the
distance which muscle's molecular motors must bridge in order to bind
and generate force \cite{Williams2013}. This distance changes during
contraction and thus regulates the force produced.

Images generated during experiments share several key features, which
serve as challenges or fiduciary marks during analysis. As seen in
Figure~\ref{fig:imgs}, the brightest part of the image background is
occluded by a circular stop. This physical block prevents damage to
the detector from the high photon flux at the center of the X-ray
beam. Surrounding the blocked region, the remainder of the image
displays an exponentially decaying background. We must locate and
model the symmetric pairs of diffraction peaks interrupting the
exponential background. Our core data analysis pipeline includes two
steps: image segmentation and image modeling with MCMC processes.

@@

The system first identifies the dark central circular blocked region
to act as a relative landmark for subsequent operations. Consistent
with experimental design, we assume it contains the center of the
diffraction pattern. The edge of the background surrounding the block
is the brightest region, so the system first splits the image between
areas with values less than and greater than two standard deviations
above the mean. This partitioning yields a binary image where the
center blocked region is surrounded by a halo of the upper end of the
pattern background and occasional dots where diffraction peaks rise
more than two standard deviations above background. We convert this
binary image to a hierarchical contour set with OpenCV. We then take
the blocked region to be the inner-most contour and model it as the
smallest enclosing circle, shown as dark magenta in
Figure~\ref{fig:workflow}.

With the central blocked region located, we identify local maxima in
three by three groups after Gaussian smoothing with kernel having a
three pixel standard deviation. We reject resulting maxima in regions
unlikely to provide peaks of interest before attempting to match peak
pairs. Masked rejection-regions consist of: 1) a circular zone around
the central blocked region where the blocking generates non-peak local
maxima, 2) areas below the 80th percentile where detector noise
dominates, 3) areas near the image edge where peaks are partially
cropped. The resulting unmasked area from which we keep maxima is
depicted as light purple overlay in Figure~\ref{fig:workflow}.

@@

Next, we cluster maxima into peak pairs based on their distance and
angle from the center of the blocked region. Starting with those
maxima nearest the blocked region, a corresponding maximum is sought
an equal distance away from the blocked region and located so that the
angle formed by the two maxima and the center of the blocked region is
180$^\circ$. If no matching maximum exists within a 10\% tolerance,
the prior maximum is discarded as a spurious peak. With peak pairs now
identified (shown as color matched dots in Figure~\ref{fig:workflow}),
the diffraction center is identified by taking the mean location
between peak pairs and the background is fit and removed.

We subtract the background by first masking arcs encompassing peak
pairs and then fitting a double exponential to a radial profile of the
remaining image. An arc swept 12$^\circ$ out on either side of each
peak pair masks the effect of the diffraction peaks on the background
(shown as a light gray arc under the peak pairs in
Figure~\ref{fig:workflow}). We calculate a radial profile of the image
around the pattern center, omitting diffraction line regions. We fit a
double exponential function of the form $background = a+ b e^{-x c} +
d e^{-x e}$ to the radial profile. From these parameters we generate
an estimated background image and subtract it from the real image,
allowing us to extract the $d{10}$ peaks as regions of interest (ROIs)
unhindered by an overlaid diffraction background.

@@

With the background subtracted and the $d_{10}$ peaks identified and
isolated from the rest of the image as ROIs, we apply Markov chain
Monte Carlo (MCMC) sampling to determine the probability distributions
from which the peak parameters could be drawn. We treat the peaks as
being drawn from an underlying Pearson VII distribution, commonly used
to fit X-ray diffraction peaks. This process allows us to generate
possible peak matches using five parameters: peak center x-location,
peak center y-location, peak height, peak spread, and peak decay rate.
We perform an initial peak fitting by residual minimization between a
generated peak and the extracted ROI. This gives a set of starting
parameters that we use, with random variation, to initialize the
positions of the MCMC agents that will explore the model space.

@@

We apply our workflow to a test corpus of 1,220 images generated using
X-ray diffraction during insect flight muscle research at Argonne
National Laboratory's BioCAT Beamline. Sample high-quality and
challenging images are shown in Figures~\ref{fig:imgs}A and
\ref{fig:imgs}B.  Our initial image segmentation step successfully
identifies the center blocked region of each image in greater than
99\% of images in our test corpus. The overall process allows us to
calculate peak-to-peak distances to sub-pixel accuracy with a
confidence interval of 90\%. MCMC sampling combined with image
segmentation allows us to precisely, accurately, and automatically
locate the $d_{10}$ peak centers and thus calculate the lattice
spacing measured by a diffraction image to within 0.03 nm.

@@

The increasing prevalence of advanced diffractive imaging techniques
demands data analytic methods that can handle, objectively and
reproducible, significant increases in the volume of data generated
and analyzed. This toolchain is a first step in a process that will
greatly increase the power and capabilities of a widely used approach
to understanding protein structure in living tissue.
